Which are the most harmful  Weather Events for public health and economy.
========================================================

# Synopsis
The goal of current data analysis is to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and answer the following questions:

1 Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2 Across the United States, which types of events have the greatest economic consequences?

First , the data was processed and cleaned,removed unnecessary rows and calculated crop and property damage.

Analysis showed that Tornados,Flood, Excessive Heat have the highest impact on populations health and economy.
## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size: [Storm data](https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2).

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.
[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata/peer2_doc/pd01016005curr.pdf)
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded

## Loading and Processing the Raw Data

```{r chunk1, cache=TRUE}
dataset<-read.csv("repdata_data_StormData.csv",stringsAsFactors=FALSE)
#evtab<-data.frame(dataset$EVTYPE,dataset$FATALITIES,dataset$INJURIES,dataset$PROPDMG,dataset$PROPDMGEXP,dataset$CROPDMG,dataset$CROPDMGEXP)
```

**Cleaning data set**

The aim of current section is to clean dataset.
First of all, the record "Summary" event is removed:
```{r}
dataset <- dataset[!grepl("Summary", dataset$EVTYPE),  ]
```
Next step is to culculate property and crop damage.The data is presenting as the data in  PROPDMG and CROPDMG and multitlier int he columns PROPDMGEXP and CROPDMGEXP.The data in these columns are presented like:
- H for hundreds ("2")

- K for thousands ("3")

- M for millions  ("6")

- B for billions  ("9")
 
All the marks and values, which are outside of these values, are removed and converted to the numerical values.
```{r}
exp<- c("^$|[-?+]", "[hH]", "[kK]", "[mM]", "[bB]")
now <- c("0", "2", "3", "6", "9")
for (i in 1:length(exp)) {
     dataset$PROPDMGEXP <- sub(exp[i], now[i], dataset$PROPDMGEXP) 
     dataset$CROPDMGEXP <- sub(exp[i], now[i], dataset$CROPDMGEXP)
}
dataset$PROPDMGEXP <- as.numeric(dataset$PROPDMGEXP)
dataset$CROPDMGEXP <- as.numeric(dataset$CROPDMGEXP)

var<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
dt<-dataset[var]
#culculate property and crop damage
dt["PROP"]<-0
dt["CROP"]<-0
dt$PROP <-dt$PROPDMG*(10^dt$PROPDMGEXP)
dt$CROP <-dt$CROPDMG*(10^dt$CROPDMGEXP)

var<-c("EVTYPE","FATALITIES","INJURIES","PROP","CROP")
dt<-dt[var]
```
Arregation the data over EVTYPE
```{r }
result <-aggregate(.~EVTYPE,data=dt,sum)

dim(result)
```

## Results

At current section we will answer the question, **which types of weather events are most harmful with respect to population health**
Health damage is calculated by summing number of fatalities and injuries per category
```{r}
fatal<-result[order(result$FATALITIES,decreasing=TRUE),][1:10,]
 head(fatal,10)
inj <-result[order(result$INJURIES,decreasing=TRUE),][1:10,]
head(inj,10)

```

```{r}
par(mfrow = c(1, 2))
barplot(fatal$FATALITIES,las=3, names.arg = fatal$EVTYPE, 
        main="Top 10 weather events \nwith by number of fatalities", ylab="number of fatalities",col="red")  
barplot(inj$INJURIES,las=3, names.arg = inj$EVTYPE, 
        main="Top 10 weather events \nwith by number of injuries", ylab="number of injuries",col="red")    
```

** Which types of events have the greatest economic consequences**

To answer this question, we find weather events which have the highest impact on property damage and crop damage.
```{r}

prop<-result[order(result$PROP,decreasing=TRUE),][1:10,]
 head(fatal,10)
crop <-result[order(result$CROP,decreasing=TRUE),][1:10,]
head(prop,10)

head(crop,10)
```
```{r}
par(mfrow = c(1, 2))
barplot(prop$PROP,las=3, names.arg = prop$EVTYPE, 
        main="Top 10 weather events \nwith by number of property damage", ylab="number of property damage",col="blue")  
barplot(crop$CROP,las=3, names.arg = crop$EVTYPE, 
        main="Top 10 weather events \nwith by number of crop damage", ylab="number of crop damage",col="blue")    
```
    
